Differential cultural reproduction of skill level and mental templates in Late Acheulean handaxe morphology: Archaeological and experimental insights

### 2023-01-20 manuscript/manuscript.Rmd
finished another round of revision.

---
title: "Differential cultural reproduction of skill level and mental templates in Late Acheulean handaxe morphology: Archaeological and experimental insights"
author: 
- Cheng Liu^[Department of Anthropology, Emory University, Atlanta, GA, USA; raylc1996@outlook.com]
- Nada Khreisheh^[The Ancient Technology Centre, Cranborne, Dorset, UK; nada.khreisheh@dorsetcouncil.gov.uk]
- Dietrich Stout^[Department of Anthropology, Emory University, Atlanta, GA, USA; dwstout@emory.edu]
- Justin Pargeter^[Department of Anthropology, New York University, New York, NY, USA; Rock Art Research Institute, School of Geography, Archaeology, and Environmental Studies, University of the Witwatersrand, Johannesburg, South Africa; justin.pargeter@nyu.edu]
abstract: "Despite the extensive literature focusing on Acheulean handaxes, especially the sources and meaning of their morphological variability, many aspects of this topic remain elusive. Archaeologists cite many factors that contribute to the considerable variation of handaxe morphology, including knapper skill levels and mental templates. Here we present results from a multidisciplinary study of Late Acheulean handaxe-making skill acquisition involving thirty naïve participants trained for up to 90 hours in Late Acheulean style handaxe production and three expert knappers. We compare their handaxe to the Late Acheulean handaxe assemblage from Boxgrove, UK. Through the principal component analysis of morphometric data derived from images, our study suggested that knapper skill levels and mental templates have a relatively clear manifestation in different aspects of handaxe morphology. The former relates to cross-sectional thinning (PC1), while the latter refers to handaxe elongation and pointedness (PC2). Moreover, we also evaluated the effects of training using the data from a 90-hour-long knapping skill acquisition experiment. We found that reaching the skill level of modern experts requires more training time than was permitted in this extensive and long-running training program.\\par \\textbf{Keywords:} Late Acheulean; Handaxe morphology; Boxgrove; Experimental archaeology; Skill level; Mental template; Cultural transmission"
date: "`r Sys.Date()`"
fontsize: 11pt
output: bookdown::pdf_document2
header-includes:
   - \usepackage{textcomp}
   - \usepackage{utopia}
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{float}
   - \floatplacement{figure}{H}
   - \usepackage{caption}
   - \captionsetup[figure]{font=scriptsize}
   - \captionsetup[table]{font=scriptsize}
linestretch: 1.5
link-citations: yes
linkcolor: blue
bibliography: bibliography.bib
csl: apa.csl
---

@@

The morphological variability of Acheulean handaxes has been one of the most well-studied and well-published topics in paleolithic archaeology [@white1998; @earlyhu1998; @key2019]. Despite the recurrent narrative emphasizing the homogeneity and longevity of handaxe assemblages on a global scale and the conservatism behind this phenomenon that evokes genetic explanations [@corbey2016; @corbey2020; @richerson2005; @sterelny2004], many researchers have recognized the diversity within what has been deemed as a unified Acheulean "tradition" and tried to dissect the sources and meaning of this variation [@lycett2008; @nowell2010; @nowell2002; @sharon2011; @moncel2018a; @moncel2018b; @moncel2018c]. More specifically, a complex suite of interconnecting factors [@lycett2015] have been identified to contribute to handaxe morphological variation, including but not limited to raw material variability [@sharon2008; @eren2014; @lycett2016; @mcnabb2015], percussor properties [@shipton2009], functional differences [@machin2007; @key2016; @white2018; @key2017; @lycett2008], reduction method/intensity [@shipton2009; @shipton2015], time budgets [@schillingerConsideringRoleTime2014], learning processes [@lycett2016; @kempe2012], social signaling [@kohn1999; @spikins2012], aesthetic preferences [@gowlett2021; @letensorer2006], knapper skill levels [@stout2014; @herzlinger2017; @caruana2021], and mental templates [@garcía-medrano2019; @hutchence2021; @schillinger2017]. From this extensive list, knapper skill levels and mental templates have been repeatedly mentioned and discussed in the now extensive corpus of handaxe studies, and Boxgrove handaxes have been one of the most studied assemblages from these two angles. Of particular attention here are the experimental works conducted by Stout et al. [-@stout2014] focusing on inferring knapping skill level and Garcia-Medrano et al. [-@garcía-medrano2019] identifying the mental template of the Boxgrove assemblage. Our paper combines these two perspectives and provides novel insights to the same archaeological assemblage by comparing it with experimentally made handaxes.

@@

@@

Drawing on these two lines of literature, we aim to explore the possibility of dissecting the interaction of skill level and mental template through a comparative study of an archaeological handaxe assemblage known for its remarkable high skill level, a reference handaxe collection produced by modern knapping experts, and an experimental handaxe sample produced by modern novice knappers. We generated the novice handaxe collection from a 90-hour skill acquisition experiment providing the opportunity to introduce the diachronic dimension of training time and interrogate its impact on the variables of interest. As such, we propose the following two interconnected research questions in this article: 1) Can skill level and mental templates be efficiently detected from handaxe morphometric data? Accordingly, we hypothesize that the morphometric variables showing overlap between Boxgrove and expert samples while being markedly different from novice samples reflect skill level differences, and all three group should show a similar mental template since this is the common target. 2) How does training affect novices' performance in these two aspects? Our hypothesis is that throughout the training the novice samples should become more similar to expert samples in both skill level and mental template.

@@

We use the statistical programming language R 4.1.1 [@rcoreteam2021] to conduct statistical analyses and data visualization in this study, particularly the R packages "FactoMineR" [@lê2008] and "ggstatsplot" [@patil2021]. As the initial step, simple visualization techniques such scatter plots are frequently used to explore the relationships between variables of interest. Given the number of variables involved in this study, we used principal component analysis (PCA) to reduce the dimension and identify the possible patterns in this morphometric data set, which is one of the most used techniques in similar studies [@iovita2011; @shipton2015; @stout2014; @garcía-medrano2020a; @garcía-medrano2020b; @herzlinger2017]. To detect the effect of training on novices' performance as compared with archaeological samples and handaxe made by experts, we also compare the corresponding metrics built on PCA across different training periods and across all groups using the Games-Howell nonparametric post-hoc test. Compared with other nonparametric tests frequently used in archaeological research for multiple group comparison such as Tukey's test, Games-Howell test does not rely on the assumptions of sample normality, and equal sample sizes and equal variance are not necessary conditions to perform this test. The sample size of each compared group can be as low as 6 [@games1976; @sauder2019]. Lastly, we compare the delta weight, as defined by the difference between initial nodule weight and end product weight, between these groups to understand the effect of training on reduction intensity. This study adheres to the principles of reproducibility and data transparency of archaeological research by depositing all the codes and data sets involved in an open-access online repository [@marwick2017], which are available as supplementary materials and can be accessed through the author's Github (<https://github.com/Raylc/Boxgrove-Exp>).

@@

Our study suggests that both skill level and mental template have a relatively clear manifestation in different aspects of handaxe morphology, where the former is related to cross-sectional thinning (PC1) while the latter relates to handaxe elongation and pointedness (PC2). Moreover, we also evaluated the effects of training using the data from a 90-hour long knapping skill acquisition experiment and confirmed the previous finding [-@pargeter2019] that reaching the skill level of modern experts requires more training time than was permitted in this extensive and long-running training program. In accordance with the existing literature on handaxe knapping skill [@caruana2020; @stout2014; @callahan1979], the results of PCA suggested that PC1 (cross-sectional thinning) is a robust indicator of skill level as it is a common feature shared by modern expert knapper and Boxgrove knappers. Thinning is regarded as a technique requiring a high knapping skill level because it requires one to carefully detach flakes in an invasive manner while not breaking the handaxe into several pieces, serving the purpose of achieving the desired convexity and/or volume. This procedure involves precise control of striking forces, strategic choice of platform external angle, and attentive preparation of bifacial intersection plane, all of which were part of our experimental training program [@callahan1979; @caruana2022; @stout2014; @pargeter2020; @shipton2013]. Experimental studies have also shown that the thinning stage of handaxe produce often involves the use of soft hammers, which is also supported by indirect archaeological evidence of flake attributes from Boxgrove [@roberts2009; @roberts1998: 384-394], although the validity of differentiating purcussor types (hard hammerstone, soft hammerstone, and antler hammer) based on flake attributes has been challenged by other experimental studies[@driscoll2014]. It should be noted that both our experts and novices frequently used soft hammers in the production of experimental assemblages. In the skill acquisition experiments, novice knappers were explicitly taught to switch to the soft hammer for thinning purposes, but some of them did not follow the instruction during the assessment. On the other hand, it has also been shown that hard hammers can also be used to achieve similar thinning results [@pelcin1997; @bradley1986], and the replicas produced by Bruce Bradley in our expert reference collection did not involve the use of soft hammers.

Given the dissimilarity of PC2 (elongation and pointedness) values between archaeological and experimental samples and its similarity among modern knappers, we argue that this dimension reflects different mental templates, where the Boxgrove assemblage displays an ovate shape featuring a wider tip while the experimental assemblages are characterized by a more pointed shape with a longer central axis. Our results regarding the ovate plan morphology of the Boxgrove assemblage generally supports what have been reported by Shipton and White [-@shipton2020] as well as Garcia-Medrano et al. [-@garcía-medrano2019]. The finding that the expert group has a mental template different from the Boxgrove assemblage is rather surprising since they were requested to mimic Boxgrove handaxes, a potential reason of which could be that these expert didn't have Boxgrove handaxes at hand as model during the manufacture and thus followed their vague memory of a "representative teardrop Late Acheulean handaxe."In general, this pattern may reflect a divergence of group-level aesthetic choices as expected under the theoretical framework of the communities of practice [@wenger1998], which could potentially provide an mechanistic explanation to some macro-level cultural phenomena such as regionalization [@garcía-medrano2022; @ashton2021; @davis2019; @shipton2020]. The most common form of learning in the experiment occurred in the group condition, where the instructor, as the competent group member, directed the joint enterprise through actively teaching multiple novices at the same time. Meanwhile, novices had the chance to also communicate and learn from their peers, producing a shared repertoire of artifacts and actions. Unfortunately, the handaxe data from the instructor (N. Khreisheh) are unavailable, but it should be noted that the instructor has learned how to knap and how to teach knapping from one of our expert knapper (Bruce Bradley). This cascading effect of social learning might explain why there is a shared mental template between the expert group and the novice group after training.

@@

In terms of our second research question, this study shows that training does have an immediate intervention effect (pre-training vs. post-training) in both PC1 (skill level) and PC2 (mental template). Nonetheless, once the training has been initiated, its effects across different assessments on both dimensions are rather non-significant. This finding provides a parallel line of evidence that corroborates what has been suggested in Pargeter et al. [-@pargeter2019] that 90 hours of training for handaxe making is still not enough for novices to reach the skill level as reflected in expert knappers, even considering the massive social support involved in the experiment set up including the direct and deliberate pedagogy and the simplified raw material procurement and preparation procedures. Methodologically speaking, this study also demonstrated that the pattern revealed by the multivariate analysis of morphometric data can nicely match with the expert knapper's 5 point grading scale of novices' knapping performances that takes multiple factors into consideration, including outcome, perceptual motor execution, and strategic understanding [See Table 2 of -@pargeter2019 for more details].

@@

Regarding the two research questions we proposed in the beginning, our case study suggested that 1) we can delineate the effects of skill level and mental template through the multivariate analysis of morphometric data, where the former is associated with cross-sectional thinning while the latter is reflected in elongation and pointedness; 2) On average training has an immediate effect of making novices to better understand the shared design targets, but 90 hours of training is still not enough for novice to reach the level of expertise as reflected in modern experienced knappers, let alone the Boxgrove tool makers. At a larger theoretical level it questions the distinction between social learning of design targets vs. individual learning of the skills needed to achieve them. To illustrate, a thin cross section could be part of a mental template or design target and was explicitly instructed by our expert instructor to novices, but novices cannot fully understand nor achieve this technological goal due to the constraint of skill level, making it a robust indicator of the latter. Traditionally archaeological experiments speaking to the literature of cultural evolution tend to use handaxe as a model artifact and focus on how copying errors emerge during the transmission of a fixed and static target using transmission chain design and alternative raw materials such as foam [@schillinger2017; @schillingerConsideringRoleTime2014; @schillinger2015]. This line of inquiry is generally characterized by high internal validity (causal mechanisms) but low external validity (generalizability to archaeological data). In contrast, our study unpacks the differential reproductions of two major sources of variation and reveals how the development of motor skill during learning is constraining the achievement of the socially learnt design target, through an actualistic experimental setting featuring a higher degree of external validity [@liu2022]. In the future, more robust experimental studies are needed to deepen our understanding of the relationship between skill acquisition and the morphological variability of handaxes in the proper developmental context [@högberg2018] as well as their implications for the biological and cultural evolution of the hominin lineages.

>>>>

---
title: "Differential cultural reproduction of skill level and mental templates in Late Acheulean handaxe morphology: Archaeological and experimental insights"
author: 
- Cheng Liu^[Department of Anthropology, Emory University, Atlanta, GA, USA; raylc1996@outlook.com]
- Nada Khreisheh^[The Ancient Technology Centre, Cranborne, Dorset, UK; nada.khreisheh@dorsetcouncil.gov.uk]
- Dietrich Stout^[Department of Anthropology, Emory University, Atlanta, GA, USA; dwstout@emory.edu]
- Justin Pargeter^[Department of Anthropology, New York University, New York, NY, USA; Rock Art Research Institute, School of Geography, Archaeology, and Environmental Studies, University of the Witwatersrand, Johannesburg, South Africa; justin.pargeter@nyu.edu]
abstract: "Despite the extensive literature focusing on Acheulean handaxes, especially the sources and meaning of their morphological variability, many aspects of this topic remain elusive. Archaeologists cite many factors that contribute to the considerable variation of handaxe morphology, including knapper skill levels and mental templates. Integrating these two lines of literature into a broader theoretical framework of cultural reproduction, here we present results from a multidisciplinary study of Late Acheulean handaxe-making skill acquisition involving thirty naïve participants trained for up to 90 hours in Late Acheulean style handaxe production and three expert knappers. We compare their handaxe to the Late Acheulean handaxe assemblage from Boxgrove, UK. Through the principal component analysis of morphometric data derived from images, our study suggested that knapper skill levels and mental templates have a relatively clear manifestation in different aspects of handaxe morphology. The former relates to cross-sectional thinning (PC1), while the latter refers to handaxe elongation and pointedness (PC2). Moreover, we also evaluated the effects of training on the differential cultural reproduction of these two aspects using the data from a 90-hour-long knapping skill acquisition experiment. We found that the desired shape of a handaxe can be relatively quickly picked up by novices, while reaching the skill level of modern experts requires more training time than was permitted in this extensive and long-running training program.\\par \\textbf{Keywords:} Late Acheulean; Handaxe morphology; Boxgrove; Experimental archaeology; Skill level; Mental template; Cultural transmission"
date: "`r Sys.Date()`"
fontsize: 11pt
output: bookdown::pdf_document2
header-includes:
   - \usepackage{textcomp}
   - \usepackage{utopia}
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{float}
   - \floatplacement{figure}{H}
   - \usepackage{caption}
   - \captionsetup[figure]{font=scriptsize}
   - \captionsetup[table]{font=scriptsize}
linestretch: 1.5
link-citations: yes
linkcolor: blue
bibliography: bibliography.bib
csl: apa.csl
---

@@

The morphological variability of Acheulean handaxes has been one of the most well-studied and well-published topics in paleolithic archaeology [@white1998; @earlyhu1998; @key2019]. Despite the recurrent narrative emphasizing the homogeneity and longevity of handaxe assemblages on a global scale and the conservatism behind this phenomenon that evokes genetic explanations [@corbey2016; @corbey2020; @richerson2005; @sterelny2004], many researchers have recognized the diversity within what has been deemed as a unified Acheulean "tradition" and tried to dissect the sources and meaning of this variation [@lycett2008; @nowell2010; @nowell2002; @sharon2011; @moncel2018a; @moncel2018b; @moncel2018c]. More specifically, a complex suite of interconnecting factors [@lycett2015] have been identified to contribute to handaxe morphological variation, including but not limited to raw material variability [@sharon2008; @eren2014; @lycett2016; @mcnabb2015], percussor properties [@shipton2009], functional differences [@machin2007; @key2016; @white2018; @key2017; @lycett2008], reduction method/intensity [@shipton2009; @shipton2015], time budgets [@schillingerConsideringRoleTime2014], learning processes [@lycett2016; @kempe2012], social signaling [@kohn1999; @spikins2012], aesthetic preferences [@gowlett2021; @letensorer2006], knapper skill levels [@stout2014; @herzlinger2017; @caruana2021], and mental templates [@garcía-medrano2019; @hutchence2021; @schillinger2017]. From this extensive list, knapper skill levels and mental templates have been repeatedly mentioned and discussed in the now extensive corpus of handaxe studies, and Boxgrove handaxes have been one of the most studied assemblages from these two angles. Of particular attention here are the experimental works conducted by Stout et al. [-@stout2014] focusing on inferring knapping skill level and Garcia-Medrano et al. [-@garcía-medrano2019] identifying the mental template of the Boxgrove assemblage. Our paper incorporates these two perspectives into a broader conceptual framework of cultural reproduction and provides novel insights to the same archaeological assemblage by comparing it with experimentally made handaxes.

## Mental template

@@

## Skill level

@@

## Cultural reproduction

The cultural reproduction, or cultural transmission as described in standard cultural evolutionary literature [@eerkens2005; @eerkens2007], of mental templates and skill levels makes them reach beyond individual-level practice and form a repetitive pattern that can be identified in archaeological records. Nonetheless, the abstract shape of handaxe as a mental template that is often pulled away from its original substrate has been frequently treated as the main research subject of cultural transmission experiments [@schillinger2017; @schillingerConsideringRoleTime2014; @schillinger2015], while how knapping skill as another source of variation is reproduced during the learning process and how it moderates the material manifestation of mental templates has been rarely discussed. The ignorance of the latter becomes one of motivations behind our terminological choice of "reproduction" over "transmission", where the former implies more than just the copying of an static image with information loss [@liu2022; @stout2021a]. This reframing essentially echoes the stance of extended evolutionary synthesis (EES) on inclusive inheritance that phenotypes are not inherited but reconstructed in development [@laland2015: 5], which has also received more attention recently in the domain of cultural evolution [@charbonneau2022; @strachan2021].

Centering around the concept of cultural reproduction, we aim to explore the possibility of dissecting the interaction of skill level and mental template through a comparative study of an archaeological handaxe assemblage known for its remarkable high skill level, a reference handaxe collection produced by modern knapping experts, and an experimental handaxe sample produced by modern novice knappers. We generated the novice handaxe collection from a 90-hour skill acquisition experiment providing the opportunity to introduce the diachronic dimension of training time and interrogate its impact on the variables of interest. As such, we propose the following two interconnected research questions in this article: 1) Can skill level and mental templates be efficiently detected from handaxe morphometric data? Accordingly, we predict that the morphometric variables showing overlap between Boxgrove and expert samples while being markedly different from novice samples reflect skill level differences, and all three group should show a similar mental template since this is a common target. 2) Does training has a differential effect in terms of the reproduction of skill level and mental templates among novices? Our hypothesis is that throughout the training the novice samples should become more similar to expert samples in both skill level and mental template, but the acquisition of the former aspect will be more challenging and thereby slower than the latter aspect. This hypothesis is informed by the previous study of Pargeter et al. [-@pargeter2020] showing that in handaxe manufacture novices' predictions of the contour of flakes to be removed are highly similar to those of expert knappers, while novices do not have the right forces and accuracy to successfully remove their target flakes to produce a nice handaxe.

@@

We use the statistical programming language R 4.1.1 [@rcoreteam2021] to conduct statistical analyses and data visualization in this study, particularly the R packages "FactoMineR" [@lê2008] and "ggstatsplot" [@patil2021]. As the initial step, simple visualization techniques such scatter plots are frequently used to explore the relationships between variables of interest. Given the number of variables involved in this study, we used principal component analysis (PCA) to reduce the dimension and identify the possible patterns in this morphometric data set, which is one of the most used techniques in similar studies [@iovita2011; @shipton2015; @stout2014; @garcía-medrano2020a; @garcía-medrano2020b; @herzlinger2017]. To detect the effect of training on novices' performance as compared with archaeological samples and handaxe made by experts, we also compare the corresponding metrics built on PCA across different training periods and across all groups using the Games-Howell nonparametric post-hoc test. Compared with other nonparametric tests frequently used in archaeological research for multiple group comparison such as Tukey's test, Games-Howell test does not rely on the assumptions of sample normality, and equal sample sizes and equal variance are not necessary conditions to perform this test. The sample size of each compared group can be as low as 6 [@games1976; @sauder2019]. Lastly, we compare the delta weight, as defined by the difference between initial nodule weight and end product weight, between these groups to understand the effect of reduction intensity on morphological variation. This study adheres to the principles of reproducibility and data transparency of archaeological research by depositing all the codes and data sets involved in an open-access online repository [@marwick2017], which are available as supplementary materials and can be accessed through the author's Github (<https://github.com/Raylc/Boxgrove-Exp>).

@@

Our study suggests that both skill level and mental template have a relatively clear manifestation in different aspects of handaxe morphology, where the former is related to cross-sectional thinning (PC1) while the latter relates to handaxe elongation and pointedness (PC2). Moreover, we also evaluated the differential effects of training using the data from a 90-hour long knapping skill acquisition experiment and confirmed the previous finding [-@pargeter2019] that reaching the skill level of modern experts requires more training time than was permitted in this extensive and long-running training program. In accordance with the existing literature on handaxe knapping skill [@caruana2020; @stout2014; @callahan1979], the results of PCA suggested that PC1 (cross-sectional thinning) is a robust indicator of skill level as it is a common feature shared by modern expert knapper and Boxgrove knappers. Thinning is regarded as a technique requiring a high knapping skill level because it requires one to carefully detach flakes in an invasive manner while not breaking the handaxe into several pieces, serving the purpose of achieving the desired convexity and/or volume. This procedure involves precise control of striking forces, strategic choice of platform external angle, and attentive preparation of bifacial intersection plane, all of which were part of our experimental training program [@callahan1979; @caruana2022; @stout2014; @pargeter2020; @shipton2013]. Experimental studies have also shown that the thinning stage of handaxe produce often involves the use of soft hammers, which is also supported by indirect archaeological evidence of flake attributes from Boxgrove [@roberts2009; @roberts1998: 384-394], although the validity of differentiating purcussor types (hard hammerstone, soft hammerstone, and antler hammer) based on flake attributes has been challenged by other experimental studies[@driscoll2014]. It should be noted that both our experts and novices frequently used soft hammers in the production of experimental assemblages. In the skill acquisition experiments, novice knappers were explicitly taught to switch to the soft hammer for thinning purposes, but some of them did not follow the instruction during the assessment. On the other hand, it has also been shown that hard hammers can also be used to achieve similar thinning results [@pelcin1997; @bradley1986], and the replicas produced by Bruce Bradley in our expert reference collection did not involve the use of soft hammers.

Given the dissimilarity of PC2 (elongation and pointedness) values between archaeological and experimental samples and its similarity among modern knappers, we argue that this dimension reflects different mental templates, where the Boxgrove assemblage displays an ovate shape featuring a wider tip while the experimental assemblages are characterized by a more pointed shape with a longer central axis. It should be noted that a thin cross section as measured by PC1 could also be part of a mental template or design target and was explicitly instructed by our expert instructor to novices, however, novices cannot fully understand nor achieve this technological goal due to the constraint of skill level, making it a robust indicator of the latter. Our results regarding the ovate plan morphology of the Boxgrove assemblage generally supports what have been reported by Shipton and White [-@shipton2020] as well as Garcia-Medrano et al. [-@garcía-medrano2019]. The finding that the expert group has a mental template different from the Boxgrove assemblage is rather surprising since they were requested to mimic Boxgrove handaxes, a potential reason of which could be that these expert didn't have Boxgrove handaxes at hand as model during the manufacture and thus followed their vague memory of a "representative teardrop Late Acheulean handaxe." In general, this pattern may reflect a divergence of group-level aesthetic choices as expected under the theoretical framework of the communities of practice [@wenger1998], which could potentially provide an mechanistic explanation to some macro-level cultural phenomena such as regionalization [@garcía-medrano2022; @ashton2021; @davis2019; @shipton2020]. The most common form of learning in the experiment occurred in the group condition, where the instructor, as the competent group member, directed the joint enterprise through actively teaching multiple novices at the same time. Meanwhile, novices had the chance to also communicate and learn from their peers, producing a shared repertoire of artifacts and actions. Unfortunately, the handaxe data from the instructor (N. Khreisheh) are unavailable, but it should be noted that the instructor has learned how to knap and how to teach knapping from one of our expert knapper (Bruce Bradley). This cascading effect of social learning might explain why there is a shared mental template between the expert group and the novice group after training.

@@

In terms of our second research question, this study shows that training does have an immediate intervention effect (pre-training vs. post-training) in both PC1 (skill level) and PC2 (mental template). Nonetheless, once the training has been initiated, its effects across different assessments on both dimensions are rather non-significant. When the performance of experts is used as a reference point here, we can see that for PC2 no significance difference is detected between early training, late training, and expert group, while for PC1 the expert group is clearly different from the training groups, supporting our hypothesis in terms of the differential cultural reproduction of mental templates and skill level. This finding provides a parallel line of evidence that corroborates what has been suggested in Pargeter et al. [-@pargeter2019] that 90 hours of training for handaxe making is still not enough for novices to reach the skill level as reflected in expert knappers, even considering the massive social support involved in the experiment set up including the direct and deliberate pedagogy and the simplified raw material procurement and preparation procedures. Methodologically speaking, this study also demonstrated that the pattern revealed by the multivariate analysis of morphometric data can nicely match with the expert knapper's 5 point grading scale of novices' knapping performances that takes multiple factors into consideration, including outcome, perceptual motor execution, and strategic understanding [See Table 2 of -@pargeter2019 for more details].

@@

Regarding the two research questions we proposed in the beginning, our case study suggested that 1) we can delineate the effects of skill level and mental template through the multivariate analysis of morphometric data, where the former is associated with cross-sectional thinning while the latter is reflected in elongation and pointedness; 2) On average training has an immediate effect of making novices to better understand the shared design targets, but 90 hours of training is still not enough for novice to reach the level of expertise as reflected in modern experienced knappers, let alone the Boxgrove tool makers, which supports our differential cultural reproduction hypothesis. At a larger theoretical level it questions the distinction between social learning of design targets vs. individual learning of the skills needed to achieve them. Traditionally archaeological experiments speaking to the literature of cultural evolution tend to use handaxe as a model artifact and focus on how copying errors emerge during the transmission of a fixed and static target using transmission chain design and alternative raw materials such as foam [@schillinger2017; @schillingerConsideringRoleTime2014; @schillinger2015]. This line of inquiry is generally characterized by high internal validity (causal mechanisms) but low external validity (generalizability to archaeological data). In contrast, our study unpacks the differential reproductions of two major sources of variation and reveals how the development of motor skill during learning is constraining the achievement of the socially learnt design target, through an actualistic experimental setting featuring a higher degree of external validity [@liu2022]. In the future, more robust experimental studies are needed to deepen our understanding of the relationship between skill acquisition and the morphological variability of handaxes in the proper developmental context [@högberg2018; @nowell2021; @lew-levy2020] as well as their implications for the biological and cultural evolution of the hominin lineages.
